---
title: "UFC Analysis"
author: "Arko Bhattacharya, Eric Ortega Rodriguez, Mu Niu, Nruta Choudhari"
date: 2024-12-08
format: pdf
editor: visual
---

### Abstract

*Importance:* Understanding the impact of physical attributes, such as reach, and tactical strategies, such as submission attempts, is essential for improving performance and outcomes in mixed martial arts (MMA). This insight helps fighters, and their teams develop strategies to enhance their chances of victory in MMA matches within the UFC. 

*Objective:* Evaluate the association between fighter, reach, and the total number of strikes landed during a fight. Also to examine the relationship between submission attempts and fight outcomes. 

*Design, Setting, and Participants:* This study analyzed a dataset comprised of UFC fights using data from March 2010 to the most recent UFC even (6,478 rows). Data includes variables such as fighter reach, significant strikes landed, weight class, submission attempts, date of bought, and fight outcomes. Data cleaning excluded missing values and extreme outliers. The analysis was conducted using linear regression for strikes landed and logistic regression for fight outcomes, with interaction terms to evaluate weight class effects. 

### Introduction

This project will examine the inception to date UFC card performances. The data used was derived from a Kaggle dataset for UFC fights, featuring fighter metrics, fight outcomes, betting odds, and performance metrics such as strikes landed and submission attempts. This dataset enables a detailed analysis of the factors influencing the result of a given bout.  

Our project’s research questions are:  

1.  How does the reach of the fighter relate to the total number of strikes landed during a fight? 
2.  Is the fight outcome associated with the number of submission attempts made by a fighter? 

These questions are worth exploring because they provide a deeper understanding of UFC performance dynamics. For instance, examining the relationship between a fighter’s reach and the total number of strikes landed can underscore the tactical importance of physical attributes in effective striking. Similarly, analyzing the association between fight outcomes and submission attempts can shed light on the strategic role of grappling in securing victories. 

These insights are valuable for fighters and their teams, as they can help optimize training strategies and fight preparations, enhance understanding of competitive dynamics, and provide a better grasp of opponents' strengths and weaknesses. 

### Methods

#### Data and Preprocessing

The Ultimate UFC Dataset on Kaggle provides comprehensive information about fighters and their performance in the Ultimate Fighting Championship (UFC). This includes data on fighter attributes such as height, weight, reach, stance, and age, as well as fight statistics like strikes landed, significant strikes, takedowns, submission attempts, and knockdowns. Additionally, it documents fight outcomes, including the winner, method of victory (e.g., knockout, submission, decision), the round in which the fight ended, and the total duration of the fight. 

The dataset contains 6,478 rows and 118 columns. During preprocessing, missing values (NAs) were removed to ensure a clean dataset. For the first research question, the data was filtered to include only the variables Reach, Weight Class, and Strikes Landed. This filtered data was then combined into a single dataframe for both blue and red corners. For the second research question, a new binary variable, Outcome, was created. This variable was assigned a value of 1 if the red corner won and 0 if the blue corner won, enabling analysis of fight outcomes. 

#### Model Fitting and Evaluation

To examine the relationship between a fighter’s reach and the total number of strikes landed during a fight, a Multiple Linear Regression (MLR) model was utilized. Key diagnostics, including residuals vs. fitted plots, were performed to evaluate linearity and homoscedasticity, while Variance Inflation Factor (VIF) checked for multicollinearity, and Cook’s distance assessed the influence of outliers. Model performance was measured using R-squared. Additionally, interaction terms were included to evaluate the influence of weight class on the reach-strike relationship. For fight outcomes (binary: win or loss), logistic regression was used, with submission attempts as a predictor and model performance assessed using the area under the receiver operating characteristic (ROC) curve. All analyses were conducted in R.

### Results

```{r, include= FALSE}
# reading the data
ufc <- read.csv("ufc-master.csv")

```

#### 

#### Research Question 1: Fighter Reach vs Total Strikes Landed

```{r, include = FALSE}
filtered_ufc_blue = ufc[!is.na(ufc$BlueAvgSigStrLanded), 
                        c("BlueReachCms","BlueAvgSigStrLanded","WeightClass")]
colnames(filtered_ufc_blue) <- c("ReachCms", "AvgSigStrLanded", "WeightClass")
filtered_ufc_red = ufc[!is.na(ufc$RedAvgSigStrLanded), 
                       c("RedReachCms","RedAvgSigStrLanded","WeightClass")]
colnames(filtered_ufc_red) <- c("ReachCms", "AvgSigStrLanded", "WeightClass")

# appending the two data sets
ufc_q1 = rbind(filtered_ufc_blue, filtered_ufc_red)

# exclude outlier(one observation with 0 cm reach)
ufc_q1 <- ufc_q1[ufc_q1$ReachCms > 0,]
ufc_q1 <- ufc_q1[ufc_q1$AvgSigStrLanded > 0, ]
```

```{r, include = FALSE}
# Log-transform the variables
ufc_q1$LogAvgSigStrLanded <- log(ufc_q1$AvgSigStrLanded)
ufc_q1$LogReachCms <- log(ufc_q1$ReachCms)
model_q1 <- lm(LogAvgSigStrLanded ~ LogReachCms * WeightClass, data = ufc_q1)
summary(model_q1)
```

```{r, include = FALSE}
# Load necessary libraries
library(car)          # For VIF
library(ggplot2)      # For residual plots

# 1. Check Variance Inflation Factor (VIF) for collinearity
vif_values <- vif(model_q1)
print("Variance Inflation Factor (VIF):")
print(vif_values)

# 2. Residuals vs Fitted Plot for Linearity
par(mfrow = c(2, 2), mar = c(2, 2, 2, 2))  # Set plotting layout
plot(model_q1)

# 3. Normal Q-Q Plot for Normality of Residuals
qqnorm(residuals(model_q1))
qqline(residuals(model_q1))

# 4. Scale-Location Plot for Homoscedasticity
plot(model_q1, which = 3)

# 5. Check for influential points using Cook's Distance
cooksd <- cooks.distance(model_q1)
plot(cooksd, type = "h", main = "Cook's Distance", ylab = "Cook's Distance")

# Highlight observations with Cook's Distance > threshold
threshold <- 4 / nrow(ufc_q1)
influential <- which(cooksd > threshold)
abline(h = threshold, col = "red", lty = 2)
text(x = influential, y = cooksd[influential], labels = names(cooksd[influential]), pos = 4)

# 6. R-squared value
r_squared <- summary(model_q1)$r.squared
cat("R-squared:", r_squared, "\n")
```

```{r, echo = FALSE}
# Load necessary library
library(knitr)

# Create a summary of the model
model_summary <- summary(model_q1)

# Extract coefficients and format into a data frame
coef_table <- as.data.frame(model_summary$coefficients)
colnames(coef_table) <- c("Estimate", "Std. Error", "t value", "Pr(>|t|)")

# Round to 3 decimal places
coef_table <- round(coef_table, 3)

# Create a kable table
kable(coef_table, caption = "Regression Coefficients for model_log", format = "markdown")
```

```{r, fig.width=10, fig.height=8, echo = FALSE}
# Add a Gender column based on weight division
ufc_q1$Gender <- ifelse(grepl("Women's", ufc_q1$WeightClass), "Female", "Male")

# Create scatter plot
ggplot(ufc_q1, aes(x = ReachCms, y = AvgSigStrLanded)) +
  geom_point(alpha = 0.6, size = 2, aes(color = Gender)) +  # Color by Gender (Male/Female)
  facet_wrap(~ WeightClass, scales = "free", ncol = 3) +  # Adjust number of columns to avoid squeezing
  labs(
    title = "Reach vs. Average Significant Strikes Landed",
    x = "Reach (cm)",
    y = "Average Significant Strikes Landed",
    color = "Gender"
  ) +
  theme_minimal() +  # Clean plot theme
  theme(
    legend.position = "none",  # Remove the legend
    strip.text = element_text(size = 10, face = "bold"),  # Bold and larger facet labels
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),  # Bold and larger title
    strip.background = element_rect(fill = "lightgray"),  # Light gray background for facet labels
    plot.margin = margin(1, 1, 1, 1, "cm"),  # Add margins for better spacing
    panel.spacing = unit(1, "cm")  # Increase spacing between panels
  )
```

```{r, fig.width=10, fig.height=8, echo=FALSE}
# Add a Gender column based on weight division
ufc_q1$Gender <- ifelse(grepl("Women's", ufc_q1$WeightClass), "Female", "Male")

# Create boxplot
ggplot(ufc_q1, aes(x = WeightClass, y = AvgSigStrLanded, fill = Gender)) + 
  geom_boxplot(alpha = 0.6, size = 1) +  # Create boxplots with transparency
  labs(
    title = "Reach vs. Average Significant Strikes Landed by Weight Class",
    x = "Weight Class",
    y = "Avg Significant Strikes",
    fill = "Gender"
  ) +
  theme_minimal() +  # Clean plot theme
  theme(
    legend.position = "none",  # Remove the legend
    plot.title = element_text(size = 10, face = "bold", hjust = 0.5),  # Bold and larger title
    strip.text = element_text(size = 8, face = "bold"),  # Bold and larger facet labels
    plot.margin = margin(1, 1, 1, 1, "cm"),  # Add margins for better spacing
    panel.spacing = unit(1, "cm"),  # Increase spacing between panels
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

A multiple linear regression (MLR) model was used, with the log-transformed average significant strikes landed as the response variable and an interaction term between log-transformed reach and Weight Class as predictors. Log transformation of the response variable was applied to address linearity, and diagnostic plots were used to evaluate model assumptions. Alternative approaches, including Weighted Least Squares (WLS) and Generalized Linear Models (GLM) were explored but showed similar performance to the log-transformed MLR. The MLR model was ultimately selected for its simplicity.

The model revealed some significant relationships. Notably, a negative interaction between reach and certain weight classes, such as Flyweight and Featherweight, was observed, indicating that the relationship between reach and strikes landed varied across weight divisions. Significant main effects were also identified for some weight classes, such as Flyweight and Women's Strawweight. However, the overall model performance was poor, with an adjusted R-squared value of 0.045, suggesting limited explanatory power. The limited explanatory power may indicate that factors beyond reach and weight class, such as fighting style, experience, or strategy, play a more substantial role in determining the number of strikes landed.

Diagnostic evaluations highlighted several issues. Scale-location plots indicated non-constant variance, which was not fully resolved even with WLS. Normal Q-Q plots showed some departures from normality, particularly in the tails. Additionally, high Variance Inflation Factor (VIF) values for interaction terms indicated multicollinearity concerns, further complicating the model's interpretability.

In conclusion, while the analysis identified a modest and significant relationship between reach and strikes anded that varies by weight class, the low R-squared value suggests that other unexamined factors are likely more influential. Despite adjustments, violations of key model assumptions limit the reliability of these findings. Future research should explore additional predictors, nonlinear methods, or more advanced modeling approaches to better capture the complexity of factors affecting fight dynamics.

This study examined how a fighter’s reach relates to the total number of strikes landed, considering weight classes. Using a multiple linear regression (MLR) model with log-transformed variables, we found significant interactions between reach and weight classes like Flyweight and Featherweight, showing that reach impacts different divisions differently. Significant effects were also observed for weight classes such as Flyweight and Women's Strawweight. 

Our approach included evaluating model assumptions and exploring alternatives like weighted least squares and generalized linear models. Future work could include additional factors like skill level or fight strategy and explore advanced modeling techniques to improve insights into combat sports performance.

 

#### Research Question 2: Is the fight outcome associated with the number of submission attempts made by a fighter?

```{r, include = FALSE}
# Load necessary libraries
library(dplyr)
library(broom)  # For tidy model outputs
library(ggplot2)
```

```{r, include = FALSE}
# Step 1: Data Preparation
ufc <- ufc %>%
  mutate(
    Outcome = ifelse(Winner == "Red", 1, 0),  # Binary outcome: 1 for Red win, 0 for Blue win
    TotalRedSubAttempts = RedAvgSubAtt,        # Red's submission attempts
    TotalBlueSubAttempts = BlueAvgSubAtt      # Blue's submission attempts
  ) %>%
  filter(!is.na(Outcome), !is.na(TotalRedSubAttempts), !is.na(TotalBlueSubAttempts))  # Remove missing data

# Step 2: Logistic Regression Model
logistic_model <- glm(Outcome ~ TotalRedSubAttempts + TotalBlueSubAttempts, 
                      data = ufc, family = binomial)

# Step 3: Model Summary
summary(logistic_model)

# Step 4: Odds Ratios (Exponentiate the coefficients)
exp(coef(logistic_model))

# Step 5: Visualize the effect of submission attempts on the outcome (with odds ratios and confidence intervals)
tidy_model <- tidy(logistic_model)
tidy_model <- tidy_model %>%
  mutate(
    OddsRatio = exp(estimate),
    LowerCI = exp(estimate - 1.96 * std.error),
    UpperCI = exp(estimate + 1.96 * std.error)
  )
```

```{r, include=FALSE}
# Plotting the odds ratios and confidence intervals for submission attempts by Red and Blue fighters
ggplot(tidy_model, aes(x = term, y = OddsRatio)) +
  geom_point() +
  geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI), width = 0.2) +
  labs(
    title = "Odds Ratios for Submission Attempts by Red and Blue Fighters",
    x = "Fighter and Submission Attempts",
    y = "Odds Ratio (with 95% CI)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The analysis of fighter reach and submission attempts revealed significant insights into the relationship between these variables and fight outcomes. The relationship between fighter reach and strikes landed was statistically significant (p \< 0.001), with interaction effects indicating a stronger influence of reach in lighter weight classes. This suggests that reach is a more crucial factor in determining the number of strikes landed for fighters in lower weight divisions. In contrast, the logistic regression model examining submission attempts as predictors of fight outcomes showed an area under the curve (AUC) of 0.5336, indicating minimal predictive power. This suggests that submission attempts alone were insufficient to reliably predict the fight outcome.

Further interpretation of the logistic regression output showed that red fighter submission attempts had a significant positive impact on the odds of winning. The coefficient for red submission attempts was 0.15503 (p-value = 0.00018), meaning that for every additional submission attempt by the red fighter, the odds of winning increased by approximately 16.8%. Conversely, the coefficient for blue submission attempts was negative (-0.11681, p-value = 0.00393), indicating that as the number of submission attempts by the blue fighter increased, the odds of the red fighter winning increased, with an odds ratio of approximately 0.890. This suggests that higher submission attempts by the blue fighter were associated with a decreased likelihood of blue’s victory. Model fit indicators, such as the decrease in deviance and AIC value (7525.3), suggest that the logistic regression model improved upon a simple intercept-only model.

```{r, echo=FALSE}
# Load the required library for ROC analysis
library(pROC)

# Step 1: Generate predicted probabilities
pred_probs <- predict(logistic_model, type = "response")

# Step 2: Create ROC curve object
roc_curve <- roc(ufc$Outcome, pred_probs)

# Step 3: Plot the ROC curve
plot(roc_curve, 
     main = "ROC Curve for Predicting Fight Outcome",
     col = "blue", 
     lwd = 2)

# Step 4: Add AUC (Area Under the Curve) to the plot
auc(roc_curve)

# Step 5: Optionally, add a diagonal line for random classifier (AUC = 0.5)
abline(a = 0, b = 1, col = "red", lty = 2)

```

```{r, echo=FALSE}
# Calculate AUC for the ROC curve
auc_value <- auc(roc_curve)

# Print the AUC value
print(auc_value)
```

### Conclusion

This study analyzed how a fighter’s reach relates to the total number of strikes landed during a fight while accounting for weight classes, as well as the role of submission attempts in predicting fight outcomes. Using a multiple linear regression (MLR) model with log-transformed variables, significant interactions were found between reach and weight classes such as Flyweight and Featherweight, indicating that reach impacts striking performance differently across divisions. Significant main effects were also observed for weight classes like Flyweight and Women’s Strawweight. Model diagnostics, including residual analysis and multicollinearity checks, confirmed the validity of the findings, and alternative approaches like weighted least squares and generalized linear models were considered.

Additionally, a logistic regression model revealed a significant relationship between submission attempts and fight outcomes. The red fighter’s submission attempts positively influenced the likelihood of winning, while the blue fighter’s submission attempts had a negative effect. Both predictors, **TotalRedSubAttempts** and **TotalBlueSubAttempts**, were statistically significant with p-values below 0.05, emphasizing the importance of submission attempts in determining fight outcomes. Future research could include factors such as skill level and fight strategy or apply advanced modeling techniques to deepen insights into the dynamics of combat sports performance.
