---
title: "UFC RESEARCH QUESTION 1"
author: "Mu Niu, Nruta Choudhari"
subtitle: "Due 5th December, 2024"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

#### Main Variables

Research Question 1 - BlueReachCms, BlueAvgSigStrLanded, RedReachCms,
RedAvgSigStrLanded, WeightClass

Research Question 2 - Outcome, RedAvgSubAtt, BlueAvgSubAtt

```{r}
library(Hmisc)

# Import the CSV into R
ufc <- read.csv("ufc-master.csv")

# View the data
head(ufc)
```

```{r}
colSums(is.na(ufc)) 
```

```{r}
# removing the data which has way too many missing values

ufc = subset(ufc, select = -c(BMatchWCRank, RMatchWCRank, RWFlyweightRank, RWFeatherweightRank, RWStrawweightRank, RWBantamweightRank, RHeavyweightRank, RLightHeavyweightRank, RMiddleweightRank, RWelterweightRank, RLightweightRank, RFeatherweightRank, RBantamweightRank, RFlyweightRank, RPFPRank, BWFlyweightRank, BWFeatherweightRank, BWStrawweightRank, BWBantamweightRank, BHeavyweightRank, BLightHeavyweightRank, BMiddleweightRank, BWelterweightRank, BLightweightRank, BFeatherweightRank, BBantamweightRank, BFlyweightRank, BPFPRank))
```

```{r}
# rechecking
colSums(is.na(ufc)) 
```

#### If we remove all the NAs from the data:

```{r}
library(dplyr)
ufc_filter_all <- ufc %>%
  filter(
    !is.na(RedAvgSubAtt),
    !is.na(BlueAvgSubAtt),
    !is.na(BlueReachCms), 
    !is.na(RedReachCms), 
    !is.na(BlueAvgSigStrLanded), 
    !is.na(RedAvgSigStrLanded),
    !is.na(TotalFightTimeSecs),
    !is.na(WeightClass)
  )
nrow(ufc_filter_all)
```

### Imputation:

```{r}
# only plotting the variables that have missing values
ufc_with_na <- ufc[, colSums(is.na(ufc)) > 0]

# Perform the naclus operation on the filtered dataset
na_patterns <- naclus(ufc_with_na)

par(mar = c(3, 1, 1, 1) + 0.1, cex = 0.9) # Adjust margins and text size


# Plot the missing value patterns for variables with missing values
naplot(na_patterns, 'na per var')
```

```{r}
#par(mar = c(3, 1, 1, 1) + 0.1, cex = 0.9)
plot(na_patterns)
```

The columns RKOdds, RSubOdds, BSubOdds, BKOdds, EmptyArena have over 20%
of the rows missing. If these are not crucial to the analysis or
confounding columns, we should drop these.

```{r}
str(ufc_with_na)
```

```{r}
str(ufc)
```

They are all numerical so we can just use one method to impute

#### Imputation

1.  **pmm** (Predictive Mean Matching):
    -   **For continuous variables**.

    -   It matches the observed values of a variable to imputed values
        based on predictive modeling.
2.  **norm** (Normal Linear Regression):
    -   **For continuous variables**.

    -   Imputes missing values based on a linear regression model
        assuming a normal distribution.
3.  **logreg** (Logistic Regression):
    -   For binary/categorical variables (0/1).

    -   Uses logistic regression for imputation.
4.  **polyreg** (Multinomial Logistic Regression):
    -   For categorical variables with more than two levels.

    -   Used when imputing missing values for categorical variables with
        multiple categories.
5.  **lda** (Linear Discriminant Analysis):
    -   **For categorical variables**.

    -   Performs classification based on linear discriminant analysis.
6.  **rf** (Random Forest):
    -   **For both continuous and categorical variables**.

    -   Imputes missing values using random forest models. It’s often
        robust and handles mixed data types well.
7.  **cart** (Classification and Regression Trees):
    -   **For continuous and categorical variables**.

    -   Uses decision trees to impute missing values.
8.  **mean**:
    -   **For continuous variables**.

    -   Imputes missing values by filling with the mean of observed
        values.
9.  **median**:
    -   **For continuous variables**.

    -   Imputes missing values by filling with the median of observed
        values.

```{r}
library(mice)
library(sjlabelled)

ufc_unlabelled = unlabel(ufc) # making a separate df just in case

ufc_unlabelled <- ufc_unlabelled %>%
  select(-c(RedFighter, BlueFighter, Location, Date, Country))

ufc.imp <- mice(ufc_unlabelled, m=5, method="pmm", print=FALSE)
```

```{r}
ufc.imp$loggedEvents
```

```{r}
# Select only numeric columns
numeric_columns <- ufc_unlabelled %>%
  select_if(is.numeric)

# Calculate the correlation matrix
cor_matrix <- cor(numeric_columns, use = "complete.obs")
print(cor_matrix)
```

#### Fitting the GLM on the basis of imputed

```{r}
# Extract all imputed datasets
ufc.imp_all <- complete(ufc.imp, action = "long")  # action = "long" returns the data in long format

filtered_ufc_blue <- ufc.imp_all %>%
  filter(!is.na(BlueAvgSigStrLanded)) %>%
  select(BlueReachCms, BlueAvgSigStrLanded, WeightClass, BlueHeightCms, BlueCurrentWinStreak) %>%
  rename(ReachCms = BlueReachCms,
         AvgSigStrLanded = BlueAvgSigStrLanded,
         Height = BlueHeightCms,
         WinStreak = BlueCurrentWinStreak)


filtered_ufc_red <- ufc.imp_all %>%
  filter(!is.na(RedAvgSigStrLanded)) %>%
  select(RedReachCms, RedAvgSigStrLanded, WeightClass, RedHeightCms, RedCurrentWinStreak) %>%
  rename(ReachCms = RedReachCms,
         AvgSigStrLanded = RedAvgSigStrLanded,
         Height = RedHeightCms,
         WinStreak = RedCurrentWinStreak)

# Combine Blue and Red Fighter data
ufc_q1 <- bind_rows(filtered_ufc_blue, filtered_ufc_red)

# Exclude outliers (one observation with 0 cm reach)
ufc_q1 <- ufc_q1 %>%
  filter(ReachCms > 0, AvgSigStrLanded > 0)

# Check missing values: no missing
colSums(is.na(ufc_q1))
```

```{r}
# Create new columns with log-transformed values in the dataset
ufc_q1$LogAvgSigStrLanded <- log(ufc_q1$AvgSigStrLanded)
ufc_q1$LogReachCms <- log(ufc_q1$ReachCms)

# Fit models for each imputed dataset manually
models <- list()  # Create an empty list to store the models

for (i in 1:5) {
  # Extract the i-th imputed dataset
  imputed_data <- complete(ufc_q1, i)
  
  # Fit the model on the imputed dataset
  models[[i]] <- lm(LogAvgSigStrLanded ~ LogReachCms * WeightClass + Height + WinStreak, data = imputed_data)
}

# Check the models
models
```

#### 
