---
title: "UFC RESEARCH QUESTION 1"
author: "Mu Niu, Nruta Choudhari"
subtitle: "Due 5th December, 2024"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

#### Main Variables

Research Question 1 - BlueReachCms, BlueAvgSigStrLanded, RedReachCms,
RedAvgSigStrLanded, WeightClass

Research Question 2 - Outcome, RedAvgSubAtt, BlueAvgSubAtt

```{r}
library(Hmisc)

# Import the CSV into R
ufc <- read.csv("ufc-master.csv")

# View the data
head(ufc)
```

```{r}
colSums(is.na(ufc)) 
```

```{r}
# removing the data which has way too many missing values

ufc = subset(ufc, select = -c(BMatchWCRank, RMatchWCRank, RWFlyweightRank, RWFeatherweightRank, RWStrawweightRank, RWBantamweightRank, RHeavyweightRank, RLightHeavyweightRank, RMiddleweightRank, RWelterweightRank, RLightweightRank, RFeatherweightRank, RBantamweightRank, RFlyweightRank, RPFPRank, BWFlyweightRank, BWFeatherweightRank, BWStrawweightRank, BWBantamweightRank, BHeavyweightRank, BLightHeavyweightRank, BMiddleweightRank, BWelterweightRank, BLightweightRank, BFeatherweightRank, BBantamweightRank, BFlyweightRank, BPFPRank))
```

```{r}
# rechecking
colSums(is.na(ufc)) 
```

#### If we remove all the NAs from the data:

```{r}
library(dplyr)
ufc_filter_all <- ufc %>%
  filter(
    !is.na(RedAvgSubAtt),
    !is.na(BlueAvgSubAtt),
    !is.na(BlueReachCms), 
    !is.na(RedReachCms), 
    !is.na(BlueAvgSigStrLanded), 
    !is.na(RedAvgSigStrLanded),
    !is.na(TotalFightTimeSecs),
    !is.na(WeightClass)
  )
nrow(ufc_filter_all)
```

### Imputation:

```{r}
# only plotting the variables that have missing values
ufc_with_na <- ufc[, colSums(is.na(ufc)) > 0]

# Perform the naclus operation on the filtered dataset
na_patterns <- naclus(ufc_with_na)

par(mar = c(3, 1, 1, 1) + 0.1, cex = 0.9) # Adjust margins and text size


# Plot the missing value patterns for variables with missing values
naplot(na_patterns, 'na per var')
```

```{r}
#par(mar = c(3, 1, 1, 1) + 0.1, cex = 0.9)
plot(na_patterns)
```

The columns RKOdds, RSubOdds, BSubOdds, BKOdds, EmptyArena have over 20%
of the rows missing. If these are not crucial to the analysis or
confounding columns, we should drop these.

```{r}
str(ufc_with_na)
```

```{r}
str(ufc)
```

They are all numerical so we can just use one method to impute

#### Imputation

1.  **pmm** (Predictive Mean Matching):
    -   **For continuous variables**.

    -   It matches the observed values of a variable to imputed values
        based on predictive modeling.
2.  **norm** (Normal Linear Regression):
    -   **For continuous variables**.

    -   Imputes missing values based on a linear regression model
        assuming a normal distribution.
3.  **logreg** (Logistic Regression):
    -   For binary/categorical variables (0/1).

    -   Uses logistic regression for imputation.
4.  **polyreg** (Multinomial Logistic Regression):
    -   For categorical variables with more than two levels.

    -   Used when imputing missing values for categorical variables with
        multiple categories.
5.  **lda** (Linear Discriminant Analysis):
    -   **For categorical variables**.

    -   Performs classification based on linear discriminant analysis.
6.  **rf** (Random Forest):
    -   **For both continuous and categorical variables**.

    -   Imputes missing values using random forest models. It’s often
        robust and handles mixed data types well.
7.  **cart** (Classification and Regression Trees):
    -   **For continuous and categorical variables**.

    -   Uses decision trees to impute missing values.
8.  **mean**:
    -   **For continuous variables**.

    -   Imputes missing values by filling with the mean of observed
        values.
9.  **median**:
    -   **For continuous variables**.

    -   Imputes missing values by filling with the median of observed
        values.

Fitting the GLM on the basis of imputed

```{r}
library(dplyr)
library(mice)

# Function to process individual imputed datasets
process_imputed_data <- function(data) {
  # Filter and process Blue Fighter data
  filtered_blue <- data %>%
    filter(!is.na(BlueAvgSigStrLanded)) %>%
    select(BlueReachCms, BlueAvgSigStrLanded, WeightClass, BlueHeightCms, BlueCurrentWinStreak) %>%
    rename(ReachCms = BlueReachCms,
           AvgSigStrLanded = BlueAvgSigStrLanded,
           Height = BlueHeightCms,
           WinStreak = BlueCurrentWinStreak)
  
  # Filter and process Red Fighter data
  filtered_red <- data %>%
    filter(!is.na(RedAvgSigStrLanded)) %>%
    select(RedReachCms, RedAvgSigStrLanded, WeightClass, RedHeightCms, RedCurrentWinStreak) %>%
    rename(ReachCms = RedReachCms,
           AvgSigStrLanded = RedAvgSigStrLanded,
           Height = RedHeightCms,
           WinStreak = RedCurrentWinStreak)
  
  # Combine Blue and Red Fighter data
  combined_data <- bind_rows(filtered_blue, filtered_red) %>%
    filter(ReachCms > 0, AvgSigStrLanded > 0)  # Exclude outliers
  
  # Add log-transformed columns
  combined_data <- combined_data %>%
    mutate(LogAvgSigStrLanded = log(AvgSigStrLanded),
           LogReachCms = log(ReachCms))
  
  # Return the processed dataset
  return(combined_data)
}

# Perform imputation
ufc_unlabelled <- unlabel(ufc) %>%
  select(-c(RedFighter, BlueFighter, Location, Date, Country))  # Remove irrelevant columns

ufc.imp <- mice(ufc_unlabelled, m = 5, method = "pmm", print = FALSE)

# Initialize a list to store processed datasets
processed_datasets <- list()

# Process each imputed dataset
for (i in 1:5) {
  imputed_data <- complete(ufc.imp, action = i)  # Extract the i-th imputed dataset
  processed_datasets[[i]] <- process_imputed_data(imputed_data)
}

# Combine all processed datasets into one
all_processed <- bind_rows(processed_datasets, .id = "Imputation")

# Inspect the result
head(all_processed)

# Fit models for each imputed dataset
models <- list()

for (i in 1:5) {
  # Extract the i-th imputed and processed dataset
  imputed_data <- processed_datasets[[i]]
  
  # Fit the linear model on the processed data
  models[[i]] <- lm(LogAvgSigStrLanded ~ LogReachCms * WeightClass + Height + WinStreak, data = imputed_data)
}

# Check the fitted models
models
```

```{r}
# Pool results across imputed datasets
pooled_model <- pool(models)

# View the summary of the pooled model
summary(pooled_model)
```

#### 

#### 
