---
title: "UFC RESEARCH QUESTION 1"
author: "Mu Niu, Nruta Choudhari"
subtitle: "Due 5th December, 2024"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

**Research Question 1:**

How does the reach of the fighter relate to the total number of strikes
landed during a fight?

-   Outcome Variable (include the name/description and type of
    variable): Reach (continuous)

-   Independent Variable: Strikes landed (continuous variable)

-   Interaction Term: Weight class, allowing analysis of the interaction
    between reach and weight class on striking effectiveness.

Data Cleaning:

```{r}
# Import the CSV into R
ufc <- read.csv("ufc-master.csv")

# View the data
head(ufc)
```

```{r}
# checking for nulls
colSums(is.na(ufc))  # Returns a count of missing values for each column)  # Total number of missing values in the dataset
```

```{r}
# check missing
sum(is.na(ufc$BlueReachCms))
sum(is.na(ufc$RedReachCms))
sum(is.na(ufc$BlueAvgSigStrLanded))
sum(is.na(ufc$RedAvgSigStrLanded))
sum(is.na(ufc$WeightClass))
```

```{r}
filtered_ufc_blue = ufc[!is.na(ufc$BlueAvgSigStrLanded), 
                        c("BlueReachCms","BlueAvgSigStrLanded","WeightClass")]
colnames(filtered_ufc_blue) <- c("ReachCms", "AvgSigStrLanded", "WeightClass")
filtered_ufc_red = ufc[!is.na(ufc$RedAvgSigStrLanded), 
                       c("RedReachCms","RedAvgSigStrLanded","WeightClass")]
colnames(filtered_ufc_red) <- c("ReachCms", "AvgSigStrLanded", "WeightClass")

# appending the two data sets
filtered_ufc = rbind(filtered_ufc_blue, filtered_ufc_red)

# exclude outlier(one observation with 0 cm reach)
filtered_ufc = filtered_ufc[filtered_ufc$ReachCms > 0,]
head(filtered_ufc)
```

Model:

Multiple Linear Regression

Model formulation:

$StrikesLanded = \beta_0 + \beta_1Reach + \beta2WeightClass + \beta_3(Reach*WeightClass) + \epsilon$

```{r}
filtered_ufc$WeightClass = factor(filtered_ufc$WeightClass)

model = lm(AvgSigStrLanded ~ ReachCms * WeightClass, data = filtered_ufc)

summary(model)

```

```{r}
par(mfrow = c(2, 2))  # Arrange diagnostic plots in a grid
plot(model)           # Produces diagnostic plots: Residuals vs. Fitted, Q-Q plot, etc.

```
