---
title: "UFC Analysis - IDS 702 Final Project"
author: "Arko Bhattacharya, Eric Ortega Rodriguez, Mu Niu, Nruta Choudhari"
date: 2024-12-15
format: pdf
editor: visual
---

### Abstract

### Introduction

The Ultimate Fighting Championship (UFC) is the world's leading mixed martial arts (MMA) promotion, known for bringing together elite fighters from diverse combat sports backgrounds. Founded in 1993, the UFC has grown into a global phenomenon, hosting events worldwide that showcase athletes competing in disciplines such as boxing, wrestling, Muay Thai, Brazilian Jiujitsu, and judo. UFC fights take place in a distinct eight-sided cage, known as the Octagon, where fighters test their skills in striking, grappling, and overall strategy under a unified set of rules. The sport has evolved significantly over the years, introducing standardized weight classes, safety regulations, and scoring systems to ensure competitive fairness and fighter safety.

This project examines UFC performances using data on UFC fights from 2010 to the present (last updated in November, 2024). The data, sourced from Kaggle, includes key fighter metrics, fight outcomes, betting odds, and performance indicators such as strikes landed and submission attempts. By leveraging this dataset, we aim to analyze factors influencing fight outcomes and performances.

Our research questions are:

1.  How does the reach of the fighter relate to the total number of strikes landed during a fight?
2.  Is the fight outcome associated with the number of submission attempts made by a fighter?

These questions are worth exploring because they provide a deeper understanding of UFC performance dynamics. For instance, examining the relationship between a fighter's reach and the total number of strikes landed can underscore the tactical performance of physical attributes in effective striking. Similarly, analyzing the association between fight outcomes and submission attempts can shed light on the strategic role of grappling in securing victories.

The findings from this analysis offer valuable insights for fighters, coaches, and analysts, helping optimize training strategies, improve fight preparation, and enhance understanding of opponents' strengths and weaknesses.

### Methods

#### Data and Preprocessing

The dataset was obtained from Kaggle, a widely recognized platform for sharing datasets and data science resources. Each row of the dataset refers to an individual bout, which refers to an individual match between two fighters. This includes data on fighter attributes such as **height, weight, reach, stance, and age**, as well as fight statistics like **strikes landed, significant strikes, takedowns, submission attempts, and knockdowns**. Additionally, it documents fight outcomes, including the **winner**, **method of victory** (e.g., knockout, submission, decision), the **round in which the fight ended**, and the **total duration of the fight**.

The dataset contains 6,478 rows across 118 columns, with several variables containing missing values. During preprocessing, columns with over 6,000 missing values were dropped due to their lack of significance and the infeasibility of imputation. Other columns had a smaller proportion of missing values, and rows with missing values in key variables (e.g., strikes landed, reach, and weight class) were removed. This resulted in a final dataset with 4,895 rows. Most of the missing values were concentrated in performance metrics, such as submission attempts or specific strike statistics.

In UFC, fighters are assigned to either the red corner or the blue corner, which indicates their position in the Octagon and helps differentiate between competitors. For the first research question, the dataset was filtered to include the variables related to reach, weight class, height, strikes landed and current win streak, ensuring that key confounding variables were included. The data for fighters in the red and blue corners were combined into a single dataframe to facilitate analysis.

For the second research question, a new binary variable, Outcome was created to indicate the winner. A value of 1 was assigned if the fighter in the red corner won, and a value of 0 if the fighter in the blue corner won. The model included variables such as submission attempts, significant strikes landed, fight duration, and weight class to account for both physical attributes and performance metrics. These variables ensured a more comprehensive analysis of the factors influencing fight outcomes while addressing potential confounders.

#### Model Fitting and Evaluation

To examine the relationship between a fighter’s reach and the total number of strikes landed during a fight, a Multiple Linear Regression (MLR) model was utilized. The model included key predictors such as logarithmic transformation of reach, logarithmic transformation of height, win streak and weight class, with an interaction term between the win streak and weight class to explore potential moderating effects. Outliers and influential points were identified using Cook's distance, and there were removed to improve model robustness. Diagnostics, including residuals vs. fitted plots, were performed to assess linearity and homoscedasticity, while Variance Inflation Factor (VIF) was used to evaluate multicollinearity. Model performance was measured using the R-squared value.

For the second research question, a logistic regression model was employed to predict fight outcomes (binary: win or loss) using submission attempts, reach, significance strikes, fight duration, and weight class as predictors. The model was refined using stepwise selection to identify the most significant predictors, and diagnostics such as Cook's distance, leverage, and deviance residuals were used to detect and remove influential points. The final logistic regression model included key predictors such as logarithmic transformations of submission attemps, logarithmic transformation of reach, logarithmic transformation of significant strikes landed, and logarithmic transformation of fight duration. Model performance was evaluated using the area under the receiver operating characteristic curve (ROC curve), and diagnostic plots were generated to assess the model's fit.

All the analyses were conducted in R.

### Results

```{r, include = FALSE}
# loading all the libraries
library(dplyr)
library(car)          # For VIF
library(ggplot2)      # For residual plots
library(broom)
library(knitr)
library(tableone)
```

```{r, include = FALSE}
# reading the data
ufc <- read.csv("ufc-master.csv")
dim(ufc)
```

```{r, include = FALSE}
# DATA CLEANING CELL
# removing the data which has way too many missing values

ufc = subset(ufc, select = -c(BMatchWCRank, RMatchWCRank, RWFlyweightRank, 
                              RWFeatherweightRank, RWStrawweightRank, RWBantamweightRank, 
                              RHeavyweightRank, RLightHeavyweightRank, RMiddleweightRank, 
                              RWelterweightRank, RLightweightRank, RFeatherweightRank, 
                              RBantamweightRank, RFlyweightRank, RPFPRank, BWFlyweightRank, 
                              BWFeatherweightRank, BWStrawweightRank, BWBantamweightRank, 
                              BHeavyweightRank, BLightHeavyweightRank, BMiddleweightRank, 
                              BWelterweightRank, BLightweightRank, BFeatherweightRank, 
                              BBantamweightRank, BFlyweightRank, BPFPRank))

# removing all missing value rows from the columns of interest
ufc_clean <- ufc %>%
  filter(
    !is.na(RedAvgSubAtt),
    !is.na(BlueAvgSubAtt),
    !is.na(BlueReachCms), 
    !is.na(RedReachCms), 
    !is.na(BlueAvgSigStrLanded), 
    !is.na(RedAvgSigStrLanded),
    !is.na(TotalFightTimeSecs),
    !is.na(WeightClass)
  )
nrow(ufc_clean)
```

#### Research Question 1: How does the reach of the fighter relate to the total number of strikes landed during a fight?

To understand the data distribution, we first computed summary statistics for the key variables: **Reach, Weight Class, Height, Win Streak, and Average Significant Strikes Landed**. Continuous variables are reported as means with standard deviations, and categorical variables are reported as counts and percentages.

```{r, include = FALSE}
filtered_ufc_blue <- ufc_clean[c("BlueReachCms","BlueAvgSigStrLanded","WeightClass", "BlueHeightCms", "BlueCurrentWinStreak")]
colnames(filtered_ufc_blue) <- c("ReachCms", "AvgSigStrLanded", "WeightClass", "Height", "WinStreak")
filtered_ufc_red <- ufc_clean[c("RedReachCms","RedAvgSigStrLanded","WeightClass", "RedHeightCms", "RedCurrentWinStreak")]
colnames(filtered_ufc_red) <- c("ReachCms", "AvgSigStrLanded", "WeightClass", "Height", "WinStreak")

# appending the two data sets
ufc_q1 <- rbind(filtered_ufc_blue, filtered_ufc_red)

# exclude outlier(one observation with 0 cm reach)
ufc_q1 <- ufc_q1[ufc_q1$ReachCms > 0,]
ufc_q1 <- ufc_q1[ufc_q1$AvgSigStrLanded > 0, ]
```
```{r}
# Summarize the data
summary_table <- ufc_q1 %>%
  group_by(WeightClass) %>%
  summarize(
    N = n(),
    Avg_Reach = paste0(round(mean(ReachCms, na.rm = TRUE), 1), " ± ", round(sd(ReachCms, na.rm = TRUE), 1)),
    Avg_Height = paste0(round(mean(Height, na.rm = TRUE), 1), " ± ", round(sd(Height, na.rm = TRUE), 1)),
    Avg_Strikes = paste0(round(mean(AvgSigStrLanded, na.rm = TRUE), 1), " ± ", round(sd(AvgSigStrLanded, na.rm = TRUE), 1)),
    Median_Streak = paste0(median(WinStreak, na.rm = TRUE), " [", quantile(WinStreak, 0.25, na.rm = TRUE), "-", quantile(WinStreak, 0.75, na.rm = TRUE), "]")
  )

# Render the table
kable(summary_table, caption = "Summary Statistics by Weight Class", align = "c")
```


```{r}
model_q1 <- lm(AvgSigStrLanded ~ ReachCms + WeightClass*WinStreak + Height, 
               data = ufc_q1)

cooks_d <- cooks.distance(model_q1)
influential <- which(cooks_d > (4 / nrow(ufc_q1)))
ufc_q1_clean <- ufc_q1[-influential, ]

model_q1_clean <- lm(AvgSigStrLanded ~ log(ReachCms) + WinStreak * WeightClass + log(Height), data = ufc_q1_clean)
summary(model_q1_clean)
par(mfrow = c(2, 2), mar = c(2,2,2,2))
plot(model_q1_clean)

vif(model_q1_clean)

r_squared <- summary(model_q1_clean)$r.squared
cat("R-squared:", r_squared, "\n")
```

A multiple linear regression (MLR) model was applied to examine the relationship between the average significant strikes landed and predictors including log-transformed Reach, log-transformed Height, Win Streak, and the interaction between Win Streak and Weight Class. Log transformations were performed on Reach and Height to address non-linearity and non-constant variance, as indicated by diagnostic plots in Appendix 1.

The adjusted $R^2$ value for the model was **0.068**, suggesting that the predictors explain approximately **6.8%** of the variability in average significant strikes landed.

The analysis revealed several important findings regarding the relationship between fighters' physical attributes and the number of significant strikes landed. The log-transformed Reach variable showed a negative relationship with average strikes landed ($\beta = -52.68, p < 0.001$), indicating that as reach increases, the average number of strikes landed decreases. The Weight Class variable also had significant effects for several divisions. Specifically, the fighters in the Featherweight class had a significantly higher number of strikes landed ($\beta = 3.15, p = 0.001$), while Flyweight fighters landed fewer strikes ($\beta = 5.67, p < 001$). On the other hand, Women's Flyweight fighters had a significant negative relationship with strikes landed ($\beta = -14.95, p < 0.001$). In terms of interaction effects, the model included interaction terms between Win Streak and Weight Class, but most of these were not significant. However, there was a marginally significant positive interaction observed for Women's Strawweight ($\beta = 3.88, p = 0.0008$), suggesting that an increasing win streak slightly positively impacts the number of strikes landed in this weight class. Finally, the logarithm of the height variable showed a weak, marginally significant negative relationship with strikes landed ($\beta = -15.59, p = 0.087$), suggesting that taller fighters might land fewer strikes on average, although the effect is not strong. These findings underscore the complex interplay between a fighter's physical characteristics, weight class, and performance outcomes, with reach and weight class being the most influential factors in predicting the number of strikes landed.

```{r}
# Plot the relationship between log(Reach) and AvgSigStrLanded
ggplot(ufc_q1_clean, aes(x = log(ReachCms), y = AvgSigStrLanded)) +
  geom_point(alpha = 0.4, color = "blue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Effect of Log(Reach) on Avg Significant Strikes Landed",
       x = "Log(Reach in Cms)", y = "Average Significant Strikes Landed")

# Interaction plot
ggplot(ufc_q1_clean, aes(x = WinStreak, y = AvgSigStrLanded, color = WeightClass)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Interaction Between Win Streak and Weight Class",
       x = "Win Streak", y = "Average Significant Strikes Landed") +
  theme_minimal()
interaction.plot(ufc_q1_clean$WeightClass, ufc_q1_clean$WinStreak, 
                 fitted(model_q1_clean), col = 1:6, lty = 1, 
                 main = "Predicted Strikes by Weight Class and Win Streak",
                 xlab = "Weight Class", ylab = "Predicted Strikes")
```

#### Research Question 2: Is the fight outcome associated with the number of submission attempts made by a fighter?

```{r}
ufc_q2 <- ufc_clean %>%
  mutate(
    Outcome = ifelse(Winner == "Red", 1, 0),  # Binary outcome: 1 for Red win, 0 for Blue win
    WeightClass = as.factor(WeightClass),
    TotalRedSubAttempts = RedAvgSubAtt,        # Red's submission attempts
    TotalBlueSubAttempts = BlueAvgSubAtt
  ) %>%
  mutate(
    LogRedSubAttempts = log1p(TotalRedSubAttempts),
    LogBlueSubAttempts = log1p(TotalBlueSubAttempts),
    LogBlueReach = log1p(BlueReachCms),
    LogRedReach = log1p(RedReachCms),
    LogBlueSigStr = log1p(BlueAvgSigStrLanded),
    LogRedSigStr = log1p(RedAvgSigStrLanded),
    LogFightTime = log1p(TotalFightTimeSecs)
  )

# Check dimensions of the cleaned dataset
dim(ufc_q2)
```

```{r}
sim_logistic_model <- glm(
  Outcome ~ 
    LogRedSubAttempts + 
    LogBlueSubAttempts + 
    LogBlueReach + 
    LogRedReach + 
    LogBlueSigStr + 
    LogRedSigStr + 
    LogFightTime + 
    WeightClass,
  data = ufc_q2, 
  family = binomial
)


step_model <- step(sim_logistic_model, direction = "both")

# Calculate Cook's distance and leverage
cooks_distance <- cooks.distance(step_model)
hat_values <- hatvalues(step_model)
residuals <- residuals(step_model, type = "deviance")

# Thresholds
n <- nrow(ufc_q2)
p <- length(coef(step_model)) - 1
cooks_threshold <- 4 / n
leverage_threshold <- 2 * (p + 1) / n

# Identify influential points
influential_points <- which(cooks_distance > cooks_threshold | 
                            hat_values > leverage_threshold |
                            abs(residuals) > 2)

# Remove influential points
ufc_q2_filtered <- ufc_q2[-influential_points, ]

# Refit the model
final_model <- glm(formula = Outcome ~ LogRedSubAttempts + LogBlueSubAttempts + 
    LogBlueReach + LogRedReach + LogBlueSigStr + LogRedSigStr, 
    family = binomial, data = ufc_q2_filtered)

# Model summary
summary(final_model)

par(mfrow=c(2,2), mar = c(2,2,2,2))

plot(final_model)
```

### Appendix

1.  Research Question 1

```{r}
summary(model_q1)
# 1. Check Variance Inflation Factor (VIF) for collinearity
vif_values <- vif(model_q1)
print("Variance Inflation Factor (VIF):")
print(vif_values)

# 2. Residuals vs Fitted Plot for Linearity
par(mfrow = c(2, 2), mar = c(2, 2, 2, 2))  # Set plotting layout
plot(model_q1)

r_squared <- summary(model_q1)$r.squared
cat("R-squared:", r_squared, "\n")
```

2.  Research Question 2

```{r}
# Model summary of the initial simple logistic model
summary(sim_logistic_model)

par(mfrow=c(2,2), mar = c(2,2,2,2))

plot(sim_logistic_model)

# Model summary of the step model
summary(step_model)

par(mfrow=c(2,2), mar = c(2,2,2,2))

plot(step_model)
```
